---
title: "Reproducible Research: Course Project 1"
author: "Pratulya Krishna"
output: 
    html_document:
        keep_md: TRUE
---

## 1. Load and preprocess data


### 1.1 Load data
```{r loaddata}
# Read in data from csv file
    activity <- read.csv("activity.csv", header = TRUE, na.strings = "NA", stringsAsFactors = FALSE)

```


### 1.2 Process data
```{r processdata}
# Process to convert character date and date form
    activity$date <- as.Date(activity$date)

```



## 2. Mean total number of steps taken per day


### Mean, median steps taken per day

```{r meanmediansteps}

# Build dataframe with total number of steps per day
    stepsperday <- data.frame(date = unique(activity$date), totalsteps = tapply(activity$steps, activity$date, sum, na.rm = TRUE))

    stepsperday$totalsteps <- as.numeric(stepsperday$totalsteps)

# Calculate and report mean and median steps
    meansteps <- mean(stepsperday$totalsteps)
    print(paste("Mean steps =", round(meansteps,0), sep = " "))

    mediansteps <- median(stepsperday$totalsteps)
    print(paste("Median steps =", round(mediansteps,0), sep = " "))

```

### Plot total number of steps per day - histogram

```{r stepshistogram}    
# Drawing histogram
    lgdMean <- paste("Mean =", round(meansteps,0), sep = " ")
    lgdMedian <- paste("Median =", round(mediansteps,0), sep = " ")
   
    hist(stepsperday$totalsteps, xlab = "", main = "")
    title(main = "Total Number of Steps per Day", xlab = "Total Number of Steps")
    
    abline(v = meansteps, col = "black", lty = 2, lwd = 2)
    abline(v = mediansteps, col = "black", lty = 2, lwd = 2)
    
    legend("topright", bty = "n", lty = 2, col = c("black", "black"), legend = c(lgdMean, lgdMedian))
    
    
```

The above graph is stored as figure1.jpeg in the figure directory


## 3. Average daily activity pattern

### Plot of time series activity

```{r avgdailysteps}

# Build dataframe with interval and steps by interval averaged across all days
    avgstepsperday <- data.frame(interval = unique(activity$interval), avgsteps = tapply(activity$steps, activity$interval, mean, na.rm = TRUE))


# Plot time series graph
    with(avgstepsperday, plot(x = interval, y = avgsteps, type = "l", xlab = "", ylab = ""))
    title(xlab = "5-Minute Interval", ylab = "Average Steps")
    title(main = "Average Number of Steps Across All Days by 5-Minute Interval")

```

The above graph is stored as figure2.jpeg in the figure directory


###  Report interval with highest activity

```{r}
# Most active interval
    maxstep <- avgstepsperday[which.max(avgstepsperday$avgsteps), "avgsteps"]
    
    maxstepinterval <- avgstepsperday[which.max(avgstepsperday$avgsteps), "interval"]
    
    print(paste("Across all days, the ",maxstepinterval,"the interval shows highest average number of steps ", round(maxstep,0), " steps.", sep = ""))

```

## 4. Impute missing values

###  Report number of missing values
```{r missval}

# Total number of missing values in the dataset
    nullRows <- table(is.na(activity$steps) == TRUE)

    print(paste("The number of rows with missing values are:", nullRows["TRUE"], sep = " "))


```

### Fill missing values, create new dataset

```{r fillval}

# Mean for that 5-minute interval, averaged over all days
    avgstepsperinterval <- data.frame(interval = unique(activity$interval), avgsteps = tapply(activity$steps, activity$interval, mean, na.rm = TRUE))

# Make new dataset activityfilled by replacing nulls in old dataset"
    activityfilled <- activity
    numRow <- nrow(activityfilled)
    i = 1

    while(i <= numRow){
        ifelse(is.na(activityfilled[i,"steps"]), activityfilled[i,"steps"] <- avgstepsperinterval[avgstepsperinterval$interval == activityfilled[i,"interval"] ,"avgsteps"], TRUE)
        
        i = i + 1
    }
    
    print("From oringal dataset 'activity', we created a new dataset 'activityfilled")
    
    print(paste("'activityfilled' returns", any(is.na(activityfilled$steps)), "for a null check.", sep = " "))

```


### New mean, median from dataset with imputed values

```{r meanmedianstepsNew}

# Total number of steps per day
    stepsperdayNew <- data.frame(date = unique(activityfilled$date), totalsteps = tapply(activityfilled$steps, activityfilled$date, sum))

    stepsperdayNew$totalsteps <- as.numeric(stepsperdayNew$totalsteps)

# Calculation new mean and median steps
    meanstepsNew <- mean(stepsperdayNew$totalsteps)
    print(paste("New mean steps =", round(meanstepsNew,0), "vs. old mean steps =", round(meansteps,0), sep = " "))

    medianstepsNew <- median(stepsperdayNew$totalsteps)
    print(paste("New median steps =", round(medianstepsNew,0), "vs. old median steps =", round(mediansteps,0), sep = " "))

```


### New histogram with imputed values

```{r stepshistogramNew, fig.width=8}
# Diviion of area for 2 graphs
par(mfrow = c(1,2))

# Old histogram
    lgdMean <- paste("Mean =", round(meansteps,0), sep = " ")
    lgdMedian <- paste("Median =", round(mediansteps,0), sep = " ")
   
    hist(stepsperday$totalsteps, xlab = "", main = "")
    title(main = "OLD: Total Steps per Day", xlab = "Total Number of Steps")
    
    abline(v = meansteps, col = "black", lty = 2, lwd = 2)
    abline(v = mediansteps, col = "black", lty = 2, lwd = 2)
    
    legend("topright", bty = "n", lty = 2, col = c("black", "black"), legend = c(lgdMean, lgdMedian), cex = 0.55)
    
    
# New histogram
    lgdMeanNew <- paste("Mean =", round(meanstepsNew,0), sep = " ")
    lgdMedianNew <- paste("Median =", round(medianstepsNew,0), sep = " ")
   
    hist(stepsperdayNew$totalsteps, xlab = "", main = "")
    title(main = "NEW: Total Steps per Day", xlab = "Total Number of Steps")
    
    abline(v = meanstepsNew, col = "black", lty = 1, lwd = 2)
    abline(v = medianstepsNew, col = "black", lty = 2, lwd = 2)
    
    legend("topright", bty = "n", lty = c(1,2), col = c("black", "black"), legend = c(lgdMeanNew, lgdMedianNew), cex = 0.55)
    
```

The above graph is stored as figure3.jpeg in the figure directory. 
  
We observe a change in the shape of the graphs based on the old and new data.  
There is a numeric change seen in mean while median remained more or less the same. This is because mean is effected by every value in data set while median relies more on the middle order data when arranged in ascending order.
  
  
## 5. Differences in activity patterns between weekdays and weekends

### Add new variable by type of day, create dataset for new plot
```{r daytypedata}

# Add a field to dataset with filled-in missing values
    activityfilled$daytype <- ifelse(weekdays(activityfilled$date) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), "weekday", "weekend")

# Factorize that field
    activityfilled$daytype <- as.factor(activityfilled$daytype)
        str(activityfilled)

# New dataset with mean steps by interval by type of day
    plotdata <- data.frame(interval = unique(activityfilled$interval), avgsteps = tapply(activityfilled$steps, list(activityfilled$interval, activityfilled$daytype), mean))

plotdata.weekday <- plotdata[,c("interval", "avgsteps.weekday")]
names(plotdata.weekday)[2] <- "avgsteps"
plotdata.weekday$daytype <- "weekday"

plotdata.weekend <- plotdata[,c("interval", "avgsteps.weekend")]
names(plotdata.weekend)[2] <- "avgsteps"
plotdata.weekend$daytype <- "weekend"

plotdata <- rbind(plotdata.weekday, plotdata.weekend)

```


### Create time series plot of activity by type of day (weekday/weekend)

```{r daytypeactivityplot}

# Using lattice package and using XYplot from it to get a continous scatter plot in x and y, syntax for lattice is graph_type(formula,data=)

library(lattice)

# Plot average number of steps by interval and type of day
xyplot(data = plotdata, avgsteps ~ interval | daytype, layout = c(1,2), type = "l", xlab = "Interval", ylab = "Number of steps", main = "Activity Pattern by Weekday / Weekend")

```

The above graph is stored as figure4.jpeg in the figure directory.